---
title: "Benford's Law Discussion - Module 3"
author: "Joy Paulsen"
output:
  html_document:
    df_print: paged
  html_notebook: default
---


## Data Prep

### Quick Summary

We will be applying some reasonable filters and then taking a look at the 2019 population estimates for 3 states from the U.S. Census (column POPESTIMATE2019).

In order to attempt to apply Benford's Law, we will take the first digit from the above column.

### Notes

The source data can be found at the [U.S. Census Website](https://www.census.gov/data/datasets/time-series/demo/popest/2010s-total-cities-and-towns.html#ds). They have datasets available for each U.S. state for years 2010-2019.

The datasets included in this repo are copies from the above link.

The states chosen for this study were Minnesota (my home state), Wisconsin (neighboring state), and Maryland (Johns Hopkins!).

### Preprocess

```{r}
library(data.table)
library(ggplot2)
library(stringr)


# preprocess minnesota dataset
mn_pop_data <- fread("./sub-est2019_27.csv")
mn_pop_data <- mn_pop_data[NAME != "Minnesota"]
mn_pop_data <- mn_pop_data[FUNCSTAT == "A"]
mn_pop_data <- mn_pop_data[POPESTIMATE2019 > 0]
mn_pop_data[, first_digit := as.numeric(
  str_sub(as.character(POPESTIMATE2019), 1, 1)
)]

# preprocess wisconsin dataset
wi_pop_data <- fread("./sub-est2019_55.csv")
wi_pop_data <- wi_pop_data[NAME != "Wisconsin"]
wi_pop_data <- wi_pop_data[FUNCSTAT == "A"]
wi_pop_data <- wi_pop_data[POPESTIMATE2019 > 0]
wi_pop_data[, first_digit := as.numeric(
  str_sub(as.character(POPESTIMATE2019), 1, 1)
)]

# preprocess maryland dataset
md_pop_data <- fread("./sub-est2019_24.csv")
md_pop_data <- md_pop_data[NAME != "Maryland"]
md_pop_data <- md_pop_data[FUNCSTAT == "A"]
md_pop_data <- md_pop_data[POPESTIMATE2019 > 0]
md_pop_data[, first_digit := as.numeric(
  str_sub(as.character(POPESTIMATE2019), 1, 1)
)]
```

### Calculate Proportions of Leading Digits

```{r}
digits <- 1:9

# minnesota benford's dataset
mn_total_rows <- mn_pop_data[, .N]
mn_proportions <- sapply(digits, function(x) {
  return(
    mn_pop_data[first_digit == x, .N] / mn_total_rows
  )
})
mn_benfords_law <- data.table(
  leading_digit = digits,
  proportion = mn_proportions,
  source = "MN Cities Population"
)

# wisonsin benford's dataset
wi_total_rows <- wi_pop_data[, .N]
wi_proportions <- sapply(digits, function(x) {
  return(
    wi_pop_data[first_digit == x, .N] / wi_total_rows
  )
})
wi_benfords_law <- data.table(
  leading_digit = digits,
  proportion = wi_proportions,
  source = "WI Cities Population"
)

# maryland benford's law dataset
md_total_rows <- md_pop_data[, .N]
md_proportions <- sapply(digits, function(x) {
  return(
    md_pop_data[first_digit == x, .N] / md_total_rows
  )
})
md_benfords_law <- data.table(
  leading_digit = digits,
  proportion = md_proportions,
  source = "MD Cities Population"
)

# true benford's law proportions
true_benfords_law <- data.table(
  leading_digit = digits,
  proportion = log10(1 + (1 / digits)),
  source = "Benford's Law"
)

# combine together for plotting
benfords_law_data <- rbindlist(
  list(true_benfords_law, mn_benfords_law, wi_benfords_law, md_benfords_law), 
  use.names=TRUE
)
benfords_law_data[, leading_digit := as.factor(leading_digit)]
```

## Size of Each Dataset

```{r}
cat(paste("Number of Cities in MN Data:", mn_total_rows, "\n"))
cat(paste("Number of Cities in WI Data:", wi_total_rows, "\n"))
cat(paste("Number of Cities in MD Data:", md_total_rows, "\n"))
```


## Plot the Proportions and Compare to the True Benford's Law


```{r fig.width=10, fig.height=6}
ggplot(benfords_law_data, aes(leading_digit, proportion, color=source)) + 
  geom_point(size=3, shape=15, alpha=0.7) +
  geom_line() +
  theme_light() +
  ggtitle("Benford's Law Applied to 2019 City Population Data for\nU.S. states Minnesota, Wisconsin, Maryland") +
  xlab("Leading Digit") +
  ylab("Proportion")
  
```


<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
